thesis2024

### 2025-03-27 intro.qmd
Creating frames article added to book.

The impact of climate on human culture, resilience, and mobility is one of the key research topics of the UP in Iberia [e.g., @bicho2017; @bicho2012; @cascalheiraHunterGathererEcodynamics2013; @morin2008; @straus1991; @bradtmoller_repeated_2012-1; @schmidtRapidClimateChange2012]. Patterns of change and adaptation have been identified primarily through lithic technological analyses, leading to the definition of the major technocomplexes that characterise the UP in Iberia: Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian. These technocomplexes exhibit distinct knapping strategies, technological patterns, and raw material use, reflecting the technological reorganisation of past human groups. Additionally, changes in territory expansion, mobility, and social networks can be observed through variations in technological choices and the presence of long-distance raw materials (Bar-yosef, 2002; Zilhão, 1991; Schmidt et al., 2021; Straus, 1991; Fullola, 1979).

The Iberian Peninsula has often been described as a refugium during the coldest periods of the Late Pleistocene (Gómez and Lunt, 2006; Gonzalez-Sampériz et al., 2010; Hewitt, 2000; Jochim, 1987; Consuegra et al., 2002; Jones, 2003; Bicho et al., 2007), providing milder conditions than northern Europe and supporting stable human occupation throughout the UP. However, during abrupt climatic events such as Heinrich Events and the Last Glacial Maximum, the region experienced colder temperatures, reduced precipitation, and vegetation shifts. Even within Iberia, north/south differences have been identified: during abrupt climatic events, arboreal taxa such as oak were replaced by infertile heathlands in northern Iberia or semi-desert steppe vegetation in southern Iberia  (Ruiz Zapata et al., 2003; Fletcher et al., 2010; Fletcher and Goni, 2008: Sanchéz-Goni et al., 2000, 2008; Naughton et al., 2007, 2009; Roucoux et al., 2001; 2005; Boessenkool et al., 2001; Turon et al., 2003). This would have resulted in the expansion of dry, risk-prone environments. Instead of being uniformly hospitable, Iberia may have exhibited a "leopard coat" distribution of ecological niches (Schmidt et al., 2021; Jennings et al. 2011). These niches likely supported continuous human occupation, leading to the formation of multicomponent sites and eco-cultural niches (Yravreda et al., 2016; Haws, 2012; Cascalheira and Bicho, 2013; Banks et al., 2008; Cascalheira et al., 2017; Fortea and Jordá, 1975).

Some studies suggest a chronological correlation between climatic shifts and the emergence of major technocomplexes, implying that cultural reorganization may have been triggered by environmental changes that altered landscapes and essential resources (e.g, Bradtmoller et al., 2012; Bicho et al., 2015; Cascalheira and Bicho, 2013). These studies also indicate shifts in territory expansion/contraction, settlement types, mobility, and intergroup contacts (Schmidt et al., 2012). While cultural changes are not necessarily deterministic responses to climate fluctuations, it is evident that hunter-gatherers adapted to their landscapes. Cultural, technological, and social transformations provide key insights into how and why human groups adapted at different times.

The study of raw materials, alongside lithic technological analyses, is a powerful tool for identifying patterns of change and adaptation, particularly in landscape use and mobility. Raw materials found at archaeological sites reflect human decision-making and technological organisation (Bamforth & Bleed, 1997). These patterns provide crucial insights into how hunter-gatherers interacted with their environments and adapted to shifting conditions (Féblot-Augustine, 1993; Kuhn, 1995; Mellars, 1996). Variability in raw material use at a given site may result from factors such as mobility and procurement strategies influenced by climate (Ambrose & Lorenz, 1990; Binford & Stone, 1985; Kuhn, 1991, 2004; Gould & Saggers, 1985; McCall, 2007), settlement patterns (Kuhn, 2004; Surovell, 2009; Grove et al., 2023), demographic changes (Clark, 1980), or exchange networks and social interactions (Akerman, Fullagar, & van Gijn, 2002; Deacon, 1989; Torrence, 1986; Wurz, 1999; Whallon, 2006).

@@

A clear example is the identification of long-distance contacts between regions in Iberia, suggesting complex mobility and procurement patterns throughout the UP (e.g., Hermida et al., 2016; Corchón Rodríguez et al., 2016; Ortega, 2003; Sánchez de la Torre et al., 2023). The Côa Valley in northeastern Portugal is a particularly emblematic case, as non-local chert is present throughout the UP sequence, with potential sources located 150–250 km away in both southwestern (central Portugal) and southeastern (central Spain) directions. These patterns have been interpreted as evidence of exchange networks facilitated by seasonal movements and social interactions between groups across Iberia (Aubry et al., 2004).

@@

Vale Boi, located in southwestern Iberia, is a multi-component site spanning several UP technocomplexes, from the Gravettian (c. 32 ka cal BP) to the Magdalenian (c. 15 ka cal BP; Nuno Bicho et al., 2007, 2012, 2013; Cascalheira et al., 2017; Cascalheira & Bicho, 2013; Manne et al., 2012; Manne & Bicho, 2011). The site, considered an eco-cultural niche, provided stable climatic conditions even during colder periods, supporting continuous residential occupations that have been consistently dated (Cascalheira & Bicho, 2013; Cascalheira et al., 2017; Belmiro et al., 2021). Extensive lithic assemblages from Vale Boi have undergone technological analysis, revealing patterns of technological change and resilience (Belmiro et al., 2021; Cascalheira et al., 2013; Gibaja & Bicho, 2013; Horta et al., 2019; Marreiros et al., 2015). The site has also been identified as a centre for knowledge transfer, particularly during the Last Glacial Maximum (Cascalheira, 2013, 2019).

Despite extensive technological studies on Vale Boi’s lithic assemblages, macroscopic, microscopic, and geochemical analyses of chert remain limited (Veríssimo, 2004; Pereira et al., 2016). This gap hinders research into raw material procurement and use, mobility, and connections between southwestern Iberia and other regions.

@@

-   Chapter 2, titled "Creating Frames of Reference for Chert Exploitation during the Late Pleistocene in Southwesternmost Iberia," offers a geological characterisation of chert sources in the Algarve region, employing macroscopic and petrographic (thin section analysis) methods. It presents a reference collection and database of local and regional chert resources, which is crucial for distinguishing between local and non-local cherts when analysing an archaeological assemblage.

-   Chapter 3, titled "Within and Beyond: Chert Procurement Patterns During the Upper Palaeolithic in Southwesternmost Iberia," focuses on the analysis of chert artefacts from the UP sequence at the archaeological site of Vale Boi, using both macroscopic and petrographic (thin section analysis) methods. These artefacts are compared with geological reference collections. The chapter offers new insights into the presence of local and non-local cherts across the various UP occupations, identifies the probable sources of non-local cherts, and explores changes in the distribution of chert types over time.

>>>>

The impact of climate on human culture, resilience, and mobility is one of the key research topics of the UP in Iberia [e.g., @bicho2017; @bicho2012; @cascalheiraHunterGathererEcodynamics2013; @morin2008; @straus1991; @bradtmoller_repeated_2012-1; @schmidtRapidClimateChange2012]. Patterns of change and adaptation have been identified primarily through lithic technological analyses, leading to the definition of the major technocomplexes that characterise the UP in Iberia: Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian. These technocomplexes exhibit distinct knapping strategies, technological patterns, and raw material use, reflecting the technological reorganisation of past human groups. Additionally, changes in territory expansion, mobility, and social networks can be observed through variations in technological choices and the presence of long-distance raw materials [@bar-yosef2002; @zilhaoPaleoliticoSuperiorEstremadura1997; @schmidtRapidClimateChange2012; @straus1991; @fullolapericotIndustriasLiticasPaleolitico1979].

The Iberian Peninsula has often been described as a refugium during the coldest periods of the Late Pleistocene [@gomezRefugiaRefugiaPatterns2007a; @gonzalez-samperiz_steppes_2010-1; @hewitt_genetic_2000; @jochimLatePleistoceneRefugia1987; @consuegraMitochondrialDNAVariation2002; @jonesMobilitySettlementResource2013; @bichoEarlyUpperPaleolithic2017], providing milder conditions than northern Europe and supporting stable human occupation throughout the UP. However, during abrupt climatic events such as Heinrich Events and the Last Glacial Maximum, the region experienced colder temperatures, reduced precipitation, and vegetation shifts. Even within Iberia, north/south differences have been identified: during abrupt climatic events, arboreal taxa such as oak were replaced by infertile heathlands in northern Iberia or semi-desert steppe vegetation in southern Iberia [@zapataPalinologicalRecordHeinrich2003; @fletcherMillennialscaleVariabilityLast2010; @fletcherOrbitalSuborbitalscaleClimate2008; @naughtonWetDryClimatic2009; @sanchezgoniEuropeanClimaticResponse2000; @sanchezgoniContrastingImpactsDansgaard2008; @naughtonPresentdayLast252007; @roucouxCombinedMarineProxy2001; @boessenkoolNorthAtlanticSeasurface2001; @turonLandseaCorrelationsLast2003; @roucouxResponseNWIberian2005]. This would have resulted in the expansion of dry, risk-prone environments. Instead of being uniformly hospitable, Iberia may have exhibited a "leopard coat" distribution of ecological niches [@schmidtRapidClimateChange2012; @jennings_southern_2011]. These niches likely supported continuous human occupation, leading to the formation of multicomponent sites and eco-cultural niches [@yravedraNotDesertedpaleoecologyHuman2016; @hawsPaleolithicSocionaturalRelationships2012; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @banksHumanEcologicalNiches2008; @forteaperezCuevaMallaetesProblemas1975].

Some studies suggest a chronological correlation between climatic shifts and the emergence of major technocomplexes, implying that cultural reorganization may have been triggered by environmental changes that altered landscapes and essential resources [e.g., @bradtmoller_repeated_2012-1; @cascalheiraHunterGathererEcodynamics2013; @bicho_rapid_2017-1]. These studies also indicate shifts in territory expansion/contraction, settlement types, mobility, and intergroup contacts [@schmidtRapidClimateChange2012]. While cultural changes are not necessarily deterministic responses to climate fluctuations, it is evident that hunter-gatherers adapted to their landscapes. Cultural, technological, and social transformations provide key insights into how and why human groups adapted at different times.

The study of raw materials, alongside lithic technological analyses, is a powerful tool for identifying patterns of change and adaptation, particularly in landscape use and mobility. Raw materials found at archaeological sites reflect human decision-making and technological organisation [@bamforth_technology_1997]. These patterns provide crucial insights into how hunter-gatherers interacted with their environments and adapted to shifting conditions [@feblot-augustinsMobilityStrategiesLate1993; @kuhn_mousterian_1995; @mellars_neanderthal_1996]. Variability in raw material use at a given site may result from factors such as mobility and procurement strategies influenced by climate [@ambroseSocialEcologicalModels1990; @binford_righteous_1985; @kuhnUnpackingReductionLithic1991; @kuhnUpperPaleolithicRaw2004; @gouldLithicProcurementCentral1985; @mccallBehavioralEcologicalModels2007], settlement patterns [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009; @groveMovingFarMoving2023a], or exchange networks and social interactions [@akermanWeaponsWunanProduction2002; @wurzHowiesonsPoortBacked1999; @whallonSocialNetworksInformation2006; @torrenceProductionExchangeStone1986].

@@

A clear example is the identification of long-distance contacts between regions in Iberia, suggesting complex mobility and procurement patterns throughout the UP [e.g., @hermidaSilexNWPeninsula2016; @corchonrodriguezMobiliteTerritoiresRelations2016; @ortegaMobilitatDesplacamentsDels2003; @sanchezdelatorreNewDataChert2023]. The Côa Valley in northeastern Portugal is a particularly emblematic case, as non-local chert is present throughout the UP sequence, with potential sources located 150–250 km away in both southwestern (central Portugal) and southeastern (central Spain) directions. These patterns have been interpreted as evidence of exchange networks facilitated by seasonal movements and social interactions between groups across Iberia [@aubryFarFlintInferring2022; @aubryRawMaterialProcurement2004].

@@

Vale Boi, located in southwestern Iberia, is a multi-component site spanning several UP technocomplexes, from the Gravettian (c. 32 ka cal BP) to the Magdalenian [c. 15 ka cal BP; @bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site, considered an eco-cultural niche, provided stable climatic conditions even during colder periods, supporting continuous residential occupations that have been consistently dated [@bicho_ecodynamics_2013; @cascalheira_2017; @belmiro_gravettian-solutrean_2021]. Extensive lithic assemblages from Vale Boi have undergone technological analysis, revealing patterns of technological change and resilience [@belmiro_gravettian-solutrean_2021; @cascalheiraValeBoiAlgarve2013; @gibajabaoProvenienceTechnologyMorphology2013; @horta_role_2019; @marreirosLithicTechnologyGravettian2015]. The site has also been identified as a centre for knowledge transfer, particularly during the Last Glacial Maximum [@cascalheiraINFLUENCIAMEDITERRANICANAS2013; @cascalheiraTerritorialityOrganizationTechnology2019].

Despite extensive technological studies on Vale Boi’s lithic assemblages, macroscopic, microscopic, and geochemical analyses of chert remain limited [@verissimo_jazidas_2004; @pereira_territory_2016-1]. This gap hinders research into raw material procurement and use, mobility, and connections between southwestern Iberia and other regions.

@@

-   Chapter 2, titled "Creating Frames of Reference for Chert Exploitation during the Late Pleistocene in Southwesternmost Iberia" [@belmiroCreatingFramesReference2023] offers a geological characterisation of chert sources in the Algarve region, employing macroscopic and petrographic (thin section analysis) methods. It presents a reference collection and database of local and regional chert resources, which is crucial for distinguishing between local and non-local cherts when analysing an archaeological assemblage.

-   Chapter 3, titled "Within and Beyond: Chert Procurement Patterns During the Upper Palaeolithic in Southwesternmost Iberia" [@belmiroChertProcurementPatterns2025] focuses on the analysis of chert artefacts from the UP sequence at the archaeological site of Vale Boi, using both macroscopic and petrographic (thin section analysis) methods. These artefacts are compared with geological reference collections. The chapter offers new insights into the presence of local and non-local cherts across the various UP occupations, identifies the probable sources of non-local cherts, and explores changes in the distribution of chert types over time.
